---
title: "Hands-on Exericse 9: [Cont. from Hands-on 7] SKATER approach"
author: "William"
date: "October 15, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
---

# Last time on Hands-on Exercise 7...

### Analytical Question

In geobusiness and spatial policy, it is a common practice to delineate the market or planning area into homogeneous regions by using multivariate data.

-   planning and implementing targeted policies will be easier

-   reduce data complexity, enabling easier analysis

-   ensure equitable distribution of resources to each homogeneous region

### Packages

-   Spatial data handling

    -   **sf**, **rgdal** and **spdep**

-   Attribute data handling

    -   **tidyverse**, especially **readr**, **ggplot2** and **dplyr**

-   Choropleth mapping

    -   **tmap**

-   Multivariate data visualisation and analysis

    -   **coorplot**, **ggpubr**, and **heatmaply**

-   Cluster analysis

    -   **cluster**

    -   **ClustGeo**

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

### Data

Two data sets will be used in this study. They are:

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*) : This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.

-   *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

Both data sets are download from [Myanmar Information Management Unit (MIMU)](http://themimu.info/)

```{r}
shan_sf <- read_rds('data/rds/shan_sf.rds')
shan_ict <- read_rds('data/rds/shan_ict.rds')
shan_sf_cluster <- read_rds('../ex7/data/rds/shan_sf_cluster.rds')
```

# SKATER Approach

To derive spatially constrained cluster

### Overall

### Compute Neighbors list

**SKATER function only support sp objects such as SpatialPolygonDataFrame**

```{r}
shan_sp <- as_Spatial(shan_sf)
```

**Compute neighbours list from polygon list**

```{r}
# poly2nd() of spdep
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

**Apply coordinates to original shan_sf to extract polygons' centroids, which are used as the nodes for the future plot**

```{r}
coords <- st_coordinates(st_centroid(st_geometry(shan_sf)))
```

**Plot `shan.nb` on top of `shan_sf`**

> `shan_sf` extends further than the graph, so its plotted first, otherwise some of the areas will be clipped, because plotting area is determined by the characteristics of the first plot

```{r}
#| code-fold: true
plot(st_geometry(shan_sf), border=grey(.5))
plot(shan.nb,
    coords, 
    col="blue", 
    add=TRUE) # adds to existing plot, otherwise will be plotted separately
```

### **Computing minimum spanning tree**

**Compute edge costs**

> edge costs: distance between nodes

```{r}
# nbcosts() of spdep package
lcosts <- nbcosts(shan.nb, shan_ict)
```

> Gives pairwise dissimilarity between its values on the five variables and the values for the neighbouring observation (from the neighbour list). Basically, this is the notion of a generalised weight for a spatial weights matrix.

**Convert `shan.nb` to a list weights object by specifying `lcosts` as the weights**

```{r}
# nb2listw() of spdep package
shan.w <- nb2listw(
    shan.nb,
    lcosts,
    style = 'B' # ensures cost values arent row-standardised
)
summary(shan.w)
```

**Compute the tree using `shan.w`**

```{r}
# mstree() of spdep package
shan.mst <- mstree(shan.w)
```

**Check its class and dimension**

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

> Note how n (number of dimensions) in `shan.w` is **55** but `dim(shan.mst) = 54` , because `shan.mst` consists on n-1 edges (links) in order to traverse all the nodes

**Plot `shan.mst`**

```{r}
#| code-fold: true
plot(st_geometry(shan_sf), border = gray(.5))

plot(
    shan.mst,
    coords,
    cex.labels = .7,
    cex.circles = .005,
    add = TRUE
)
```

### Compute the spatially constrained cluster

```{r}
# skater() of spdep package.
clust6 <- skater(
    edges = shan.mst[,1:2], # exclude the 3rd col (cost)
    data = shan_ict,
    method = 'euclidean',
    ncuts = 5 # one less than number of clusters
)
```

> `clust6` is object `skater`

```{r}
str(clust6)
```

```{r}
ccs6 <- clust6$groups
table(ccs6)
```

> groups: shows number of observations in each cluster

**Plot pruned tree with 5 clusters**

```{r}
#| code-fold: true
#| warning: false
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(
    clust6, 
    coords, 
    cex.labels = .7 ,
    groups.colors=c("red","green","blue", "brown", "pink"),
    cex.circles=0.005,
    add=TRUE
    )
```

### **Visualising the clusters in choropleth map**

```{r}
# compute proximity matrix
proxmat <- dist(shan_ict, method = 'euclidean')

# agglomeration method (ward.D) to compute the hierachical cluster
hclust_ward <- hclust(proxmat, method = 'ward.D')

# retain 6 clusters
groups <- as.factor(cutree(hclust_ward, k=6))

shan_sf_cluster <- cbind(  # append groups onto shan_sf to produce an output sf object
    shan_sf, 
    as.matrix(groups)) %>% # convert 'list' object group into a matrix
  rename(`CLUSTER`=`as.matrix.groups.`)
```

```{r}
#| code-fold: true
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

```{r}
#| code-fold: true
#| warning: false
#| message: false
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

# ClustGeo package

To perform non-spatially constrained hierarchical cluster analysis and spatially constrained cluster analysis

### Overall

### **Ward-like hierarchical clustering**

```{r}
#| code-fold: true
# hclustgeo() to perform a typical Ward-like hierarchical clustering
nongeo_cluster <- hclustgeo(proxmat)
plot(
    nongeo_cluster,
    cex = .5
)
rect.hclust(
    nongeo_cluster,
    k = 6,
    border = 2:5
)
```

### Form clusters

```{r}
#| code-fold: true
groups <- as.factor(cutree(nongeo_cluster, k=6))
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

Derive Spatial Distance Matrix

```{r}
dist = st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist) # convert df to matrix
```

`choicealpha()` will be used to determine a suitable value for the mixing parameter alpha

```{r}
#| code-fold: true
#| layout-nrow: 1
# determine a suitable value for the mixing paramete

cr <- choicealpha(
    proxmat, 
    distmat, 
    range.alpha = seq(0, 1, 0.1), # start, stop, step
    K=6, 
    graph = TRUE)
```

Use alpha = 0.2

```{r}
#| code-fold: true
# perform a typical Ward-like hierarchical clustering
clustG <- hclustgeo(
    proxmat,
    distmat,
    alpha = 0.2
)

# derive cluster object
groups <- as.factor(cutree(clustG, k=6))

# join ClusterGeo groups list of formed clusters with shan_sf
shan_sf_Gcluster <- cbind(
    shan_sf,
    as.matrix(groups)) %>% 
    rename(`CLUSTER` = `as.matrix.groups.`)

# plot newly delineated spatially constrained clusters
qtm(shan_sf_Gcluster, "CLUSTER")
```

# Visual Interpretation of Clusters

### Univariate

Compute boxplots

```{r}
colnames(shan_sf_ngeo_cluster)
```

```{r}
#| code-fold: true
#| layout-nrow: 3
#| layout-ncol: 2
ggplot(
    data = shan_sf_ngeo_cluster,
    aes(x = CLUSTER, y = RADIO_PR)) +
    geom_boxplot()
ggplot(
    data = shan_sf_ngeo_cluster,
    aes(x = CLUSTER, y = LLPHONE_PR)) +
    geom_boxplot()
ggplot(
    data = shan_sf_ngeo_cluster,
    aes(x = CLUSTER, y = MPHONE_PR)) +
    geom_boxplot()
ggplot(
    data = shan_sf_ngeo_cluster,
    aes(x = CLUSTER, y = COMPUTER_PR)) +
    geom_boxplot()
ggplot(
    data = shan_sf_ngeo_cluster,
    aes(x = CLUSTER, y = INTERNET_PR)) +
    geom_boxplot()
```

### Multivariate

Parallel coordinate plot:

```{r}
#| code-fold: true
ggparcoord(
    data = shan_sf_ngeo_cluster,
    columns = c(17:21),
    scale = 'globalminmax', # no scaling, determined by global data range
    alphaLines = .2,
    boxplot = TRUE, # boxplot overlay
    title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30)
)
```

> Reveals that households in Cluster 4 townships tend to own the highest number of TV and mobile-phone. On the other hand, households in Cluster 5 tends to own the lowest of all the five ICT
>
> Also note `scale` have other scaling methods, explore and select the optimal.

### Compute the summary statistics

```{r}
#| code-fold: true

shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```
