---
title: "Take-Home Exercise 3: Prototyping Modules for Geospatial Analytics Shiny Application"
author: "William"
date: "October 16, 2024"
date-modified: "last-modified"
execute: 
  eval: false
  echo: true
  freeze: true
format: 
  html:
    code-fold: true
    code-tools: true
    code-link: true
    grid:
      margin-width: 350px
---

NOTE: EVAL: FALSE

# Overview

# Motivation

# Methodology

My group and I will be investigating crime in Malaysia. This is how we decided to split the workload of the initial analysis:

-   each of us focus on one crime subtype under 'Assault' (referred to as 'violent crimes' on the source website)

    -   will: causing injury

    -   marcus: murder

    -   santhya:

-   Clustering analysis:

    -   ESDA (LISA maps with main input parameters: crime type, contiguity?/bandwidth analysis/k-means method, confidence level)

    -   Clustering

        -   santhya: hclust
        -   marcus: clustGEO
        -   will: SKATER

### R packages used

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, ggpubr, cluster, factoextra, NbClust, heatmaply, corrplot, psych, tidyverse, GGally)
```

### Datasets

+---------------------------------------------------+-----------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Data                                              | Description                                                                 | Source                                                                                                                                                                          |
+===================================================+=============================================================================+=================================================================================================================================================================================+
| Crimes in Malaysia                                | Number of crimes in Malaysia by district, crime category, and type of crime | [data.gov.my](https://data.gov.my/data-catalogue/crime_district?state=johor&district=all&category=assault&type=all&visual=timeseries)                                           |
+---------------------------------------------------+-----------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Population figures                                | Population at state level from 1970 - 2024                                  | [data.gov.my](https://data.gov.my/data-catalogue/population_state?state=johor&visual=table)                                                                                     |
+---------------------------------------------------+-----------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Malaysia's Sub-national Administrative Boundaries | Malaysia administrative level 0-2 boundaries, where:                        | [data.humdata.org](https://data.humdata.org/dataset/b85a92b8-e29d-444c-9990-94e4a3a85480/resource/fb5bcaee-8421-4ee9-9fe4-b31a275b0ba7/download/mys_adm_unhcr_20210211_shp.zip) |
|                                                   |                                                                             |                                                                                                                                                                                 |
|                                                   | -   level 1: state                                                          |                                                                                                                                                                                 |
|                                                   |                                                                             |                                                                                                                                                                                 |
|                                                   | -   level 2: district                                                       |                                                                                                                                                                                 |
+---------------------------------------------------+-----------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

```{r}
library(readr)
crime_district <- read_csv("data/crime_district.csv")
population_state <- read_csv("data/population_state.csv")
msia_adm1_sf <- st_read(dsn = 'data/mys_adm_unhcr_20210211_shp/', layer = 'mys_admbnda_adm1_unhcr_20210211')
```

**`crime_district`** :

-   remove rows containing country-aggregated data, 'property'-type crimes

-   remove columns where `category='assault'` and `district='All'`

    ```{r}
    crime_state <- crime_district %>% 
        filter(state != "Malaysia") %>% 
        filter(category != "property") %>% 
        filter(district == "All") %>% 
        select(-c(2:3))
    ```

`population_state` :

-   remove all the dates outside the 2016 to 2022 time period

-   select rows aggregated by sex, age and ethnicity (where data under the corresponding columns equals both or overall)

    ```{r}
    {r}
    library(dplyr)
    library(lubridate)

    # population_state_filtered <- 
    pop_data <- population_state %>%
        filter(year(ymd(date)) >= 2016 & year(ymd(date)) <= 2022) %>% 
        filter(sex == 'both') %>% 
        filter(age == 'overall') %>% 
        filter(ethnicity == 'overall') %>% 
        select(-c(3:5))
    ```

::: {.callout-important appearance="minimal"}
**Population figures are in the '000s (Thousands of people)**
:::

`msia_adm1_sf` :

-   extract `ADM1_EN`, `ADM1_PCODE`, `Shape_` and `geometry` data

    ```{r}
    adm1_sf <- msia_adm1_sf %>% select(-c(3:7))
    ```

Ensure that consistent state labels across datasets.

```{r}
unique(pop_data$state)
unique(crime_state$state)
unique(adm1_sf$ADM1_EN)
```

After running each dataset through `unique()` , **W.P. Labuan** **and** **W.P. Putrajaya are missing from crime data.** This is because "data for W.P. Putrajaya and W.P. Labuan are classified under **W.P. Kuala Lumpur** and **Sabah** respectively" ([source](https://data.gov.my/data-catalogue/crime_district?state=johor&district=all&category=assault&type=all&visual=table#:~:text=One%20of%2016%20states%2C%20or%20Malaysia%3B%20however%2C%20data%20for%20W.P.%20Putrajaya%20and%20W.P.%20Labuan%20are%20subsumed%20under%20W.P.%20Kuala%20Lumpur%20and%20Sabah%20respectively.)). Hence, I will have to modify `pop_data` (specifically W.P. Kuala Lumpur data).

```{r}
fixed_pop_data <- pop_data %>%
  mutate(state = case_when( # case_when allows for multiple if-else statements
    state == "W.P. Putrajaya" ~ "W.P. Kuala Lumpur",
    state == "W.P. Labuan" ~ "Sabah",
    TRUE ~ state # assumes no NA, which there arent any
  )) %>%
  group_by(state, date) %>%
  summarise(population = sum(population), .groups = 'drop')
```

```{r}
#| echo: false
fixed_pop_data <- read_rds('data/rds/fixed_pop_data.rds')
crime_state <- read_rds('data/rds/crime_state.rds')
adm1_sf <- read_rds('data/rds/adm1_sf.rds')
```

Derive crime **rate** rather than using absolute crime figures. Benefits include:

::: column-margin
$$
\text{Crime Rate} = \Bigg( \frac{\text{Number of Crimes}}{\text{Population in Thousands}} \Bigg) \times \ 1\text{,}000
$$
:::

-   Easier to judge the prevalence of crime relative to the population size of the state

-   Understanding rates can help communicate the actual risk of crime to the community, helping to mitigate unnecessary fear

    ```{r}
    crime_rates_state <- crime_state %>% 
        left_join(fixed_pop_data, by = c("state", "date")) %>% 
        mutate(crime_rate = (crimes / population) * 1000)
    ```

    ```{r}
    #| eval: false
    write_rds(crime_rates_state, 'data/rds/crime_rates_state.rds')
    ```

### Tabs (Creating, Modifying, Exploring...)

# \[Prototype\] Shiny Application UI Storyboard

> [Storyboard Guide](https://giniceseah.netlify.app/posts/2021-07-31-storyboard/)

summary text

### Tab 1

Figure

~*Figure caption*~

Description

\<repeat\>

# Conclusion
