---
title: "ICE 2: "
author: "William"
date: "August 26, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
---

# Getting Started

```{r}
pacman::p_load(tidyverse,sf,ggstatsplot,tmap)
```

# Importing Data

### [2014 Sub-zone Data](https://beta.data.gov.sg/datasets?query=Master+Plan+2014+Subzone+Boundary+%28Web%29&resultId=d_5cb80a95445f236737f6bc2bfe5f159d) (`shp` & `kml`)

```{r }
mpsz14_shp <- st_read(dsn = 'data/geospatial/', layer = 'MP14_SUBZONE_WEB_PL')
# layer means sf will search for the engine(?)
# layer wont need extension because it knows to seek for a shapefile
#'<-' or '=' works on windows+macOS
```

> Note: R is a OOP enabled language, sf being the object class

```{r eval=FALSE}
View(mpsz14_shp)
```

```{r eval=FALSE}
mpsz14_kml = st_read('data/geospatial/MasterPlan2014SubzoneBoundaryWebKML.kml')
# supposed to trigger a error message saying its unsupported / corrupted
```

**`MasterPlan2014SubzoneBoundaryWebKML.kml` from data.gov.sg is apparently corrupted.**

**This is a quick fix:**

```{r warning=FALSE}
st_write(mpsz14_shp, 'data/geospatial/MP14_SUBZONE_WEB_PL.kml', delete_dsn = TRUE)
```

### [**2019 Subzone Boundary (No sea)**](https://data.gov.sg/collections/1749/datasets/d_0900fdcfdb4666fe2a630f37a532fc06/view) (`kml`, `shp`, `geojson`)

```{r}
mpsz19_kml = st_read('data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml')
```

```{r}
mpsz19_shp <- st_read(dsn = 'data/MPSZ-2019/', layer = 'MPSZ-2019')
```

```{r}
mpsz19_geojson <- st_read('data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson')
```

> Note: geojson format is very messy because the data is buried under html tags so lines of code is needed to extract; while in the kml / shp formats the data is readily accessible

> Note: under 'geometry' if the values tend to be large, its because its in PCS (measure in meters)

**In-class discussion**

<details>

------------------------------------------------------------------------

### [Singapore Residents by Planning Area / Subzone, Age Group, Sex and Floor Area of Residence 2023](https://www.singstat.gov.sg/-/media/files/find_data/population/statistical_tables/respopagesexfa2023e.ashx)

> Note: use the .**csv** file

```{r}
popdata2023 = read_csv('data/respopagesexfa2023.csv')
```

------------------------------------------------------------------------

</details>

### [Pre-School Location](https://data.gov.sg/datasets?query=Pre-Schools+Location&page=1&resultId=d_61eefab99958fd70e6aab17320a71f1c) (both `kml` & `geojson` )

```{r}
preschool_kml = st_read('data/PreSchoolsLocation.kml')
```

```{r}
preschool_geojson <- st_read('data/PreSchoolsLocation.geojson')
```

<details>

# Transforming Coordinate System

```{r}
mpsz19_shp_3414 = mpsz14_shp %>% # pipe operator; saves memory
  st_transform(crs = 3414)
```

```{r}
preschool_kml_3414 = preschool_kml %>% 
  st_transform(crs = 3414)
```

**Checking CRS**

```{r}
st_crs(mpsz19_shp)
mpsz19_shp_3414
```

# Data Wrangling

**Point-in-Polygon count:** count the number of pre-schools in each planning sub-zone

```{r}
mpsz19_shp_3414 <- mpsz19_shp_3414 %>% mutate(`PreSch Count` = lengths(st_intersects(mpsz19_shp_3414, preschool_kml_3414)))
```

Write a single line code to perform the following tasks:

1.  **Derive the area** of each planning sub-zone.

2.  Drop the **unit of measurement** of the area (i.e. m\^2)

3.  **Calculate the density** of pre-school at the planning sub-zone level

```{r}
mpsz19_shp_3414 <- mpsz19_shp_3414 %>% 
  mutate(
    Area = units::drop_units(st_area(.)),
    `PreSch Density` = `PreSch Count` / Area * 1000000
    )
```

# Statistical Analysis

Using Exploratory Data Analysis (EDA) and Confirmatory Data Analysis (CDA) methods to **explore and confirm the statistical relationship between Pre-school Density and Pre-school count.**

```{r}
mpsz19_shp_3414$`PreSch Density` <- as.numeric(as.character(mpsz19_shp_3414$`PreSch Density`))
mpsz19_shp_3414$`PreSch Count` <- as.numeric(as.character(mpsz19_shp_3414$`PreSch Count`))
mpsz19_shp_3414 <- as.data.frame(mpsz19_shp_3414)

ggscatterstats(
  data = mpsz19_shp_3414,
  x = `PreSch Density`,
  y = `PreSch Count`,
  type = 'parametric'
)
```

# **Prepare a data frame using the population data**

```{r}
popdata2023 = popdata2023 %>% 
  group_by(PA,SZ,AG) %>% 
  summarise(POP=sum(Pop)) %>% 
  ungroup() %>% 
  pivot_wider(names_from = AG,values_from = POP)
```

> pivot_wider() transposes the data table structure for easier function application

**Examine the data structure**

```{r}
str(popdata2023)
```

> Note: **age ranges are not in sequential order**

```{r}
colnames(popdata2023)
```

> Note: **index starts from 1**

**Fix the data table (override the previous)**

Write a code chunk to derive a tibble data.framewith the following fields PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY where by:

-   YOUNG: age group 0 to 4 until age groyup 20 to 24,

-   ECONOMY ACTIVE: age group 25-29 until age group 60-64,

-   AGED: age group 65 and above,

-   TOTAL: all age group, and

-   DEPENDENCY: the ratio between young and aged against economy active group.

```{r}
popdata2023 = popdata2023 %>% 
  # CALCULATING AGE GROUP TOTALS
  mutate(`YOUNG`=rowSums(.[3:6])+rowSums(.[14])) %>%                # CREATES new col 'YOUNG'
  mutate(`ECONOMY ACTIVE`=rowSums(.[7:13])+rowSums(.[15])) %>%   # CREATES new col `ECONOMY ACTIVE`
  mutate(`AGED`=rowSums(.[16:21])) %>%                              # CREATES new col `AGED`
  mutate(`TOTAL`=rowSums(.[3:21])) %>%                              # CREATES new col `TOTAL`
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)/`ECONOMY ACTIVE`) %>%   # CREATES new col `DEPENDENCY` to calculate dependency ratio
  select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`) # determines column order and contents in OUTPUT DATASET
```

Because **R is case sensitive**, convert into all same uppercase or lowercase in the datasets.

```{r}
popdata2023 = popdata2023 %>% 
  mutate_at(
    .vars = vars(PA,SZ),
    .funs = list(toupper)
    )
```

Perform **left-join** whereby the join fields are `SUBZONE_N` from the `mpsz19_shp` sf data.frame and `SZ` from the `popdata2023` data.frame.

```{r}
mpsz_popdata2023 = left_join(mpsz19_shp_3414,popdata2023, by = c('SUBZONE_N'='SZ'))
```

> Note: In the c() function, the column names' positions correspond to which data table is the left table

Plot the choropleth map.

```{r}
tmap_mode('plot')

tm_shape(st_sf(mpsz_popdata2023)) +
  tm_fill( # either a fixed color or a color palette is mapped to the variable
    'DEPENDENCY',
    style = 'quantile',
    palette = 'Blues',
    title = 'Dependency ratio'
  ) +
  tm_layout(
    main.title = "Distribution of Dependency Ratio by planning subzone",
    main.title.position = "center",
    main.title.size = 1,
    legend.title.size = 1,
    legend.height = 0.45, 
    legend.width = 0.35,
    bg.color = "#E4D5C9",
    frame = F
  ) +
  
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 1.5) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics (DOS)", 
             position = c("left", "bottom"))
```

</details>
