---
title: "Take-Home Exercise 1: Application of Spatial and Spatio-temporal Point Patterns Analysis to discover the geographical distribution of Armed Conflict in Myanmar"
author: "William"
date: "September 7, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
---

# Objectives

I will be using armed conflict [data](https://acleddata.com/data-export-tool/) of Myanmar from January 2021 to June 2024, aka Armed Conflict Location & Event Data (ACLED).

I will be focusing on thsese main event types:

-   Battles,

-   Explosion/Remote violence,

-   Strategic developments, and

-   Violence against civilians.

Along the way, I will be exploring the use of different functions to obtain more aesthetically pleasing outputs for fun.

# Data

## Packages

These R packages will be used:

-   **sf**: importing, managing, and processing geospatial data, and

-   **tidyverse**: performing data science tasks such as importing, wrangling and visualising data.

-   **readr**

    -   `read_csv()`: reading ACLED csv data

-   **janitor**: data cleaning

-   **lubridate**: date formatting

-   **Extra (TBD):**

    -   **janitor**: data cleaning

    -   **lubridate**: date formatting

    -   **skimr**: provide summary statistics

    -   **scrutiny**: data wrangling

    -   cowplot - add-on to ggplot for higher quality figures

    -   [htmlwidgets](https://www.htmlwidgets.org/showcase_leaflet.html) for R - interactive visualisations (could be used for visual analytics)

        -   leaflet - geo-spatial mapping

    -   

```{r}
pacman:::p_load(sf,tidyverse,janitor,skimr,lubridate,tmap) # more to be added
```

## ACLED

```{r}
acled_sf <- read_csv('data/2021-01-01-2024-07-01-Myanmar.csv')
```

## Metadata

Click to expand –\>

<details>

+----------------------+----------------------------------------------------------------------------------------------------------+
| Variable             | Description                                                                                              |
+======================+:=========================================================================================================+
| 'event_id_cnty'      | unique alphanumeric ID by number & country acronym                                                       |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'event_date'         | day-month-year of event                                                                                  |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'year'               | self-explanatory                                                                                         |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'time_precison'      | numeric code for level of precison of data records                                                       |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'disorder_type'      | i.e. 'Demonstrations','Political violence','Political violence; Demonstrations','Strategic developments' |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'event_type'         | nature of event                                                                                          |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'sub_event_type'     | provides further classification to event_type                                                            |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'actor1'             | involved actors along with their respective affiliates                                                   |
|                      |                                                                                                          |
| 'assoc_actor_1'      |                                                                                                          |
|                      |                                                                                                          |
| 'actor2'             |                                                                                                          |
|                      |                                                                                                          |
| 'assoc_actor_2'      |                                                                                                          |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'inter1'             | numeric code between 0 and 8 encoding different actors\                                                  |
|                      |                                                                                                          |
| 'inter2'             |                                                                                                          |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'interaction'        | indicates the 2 actor types interacting in the event (encoded inter1 & inter2 by concatenation)          |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'civilian_targeting' | indicates whether the event involved civilian targeting (e.g. null if false)                             |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'iso'                | unique 3-digit numeric code assigned to each country or territory according to ISO 3166                  |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'region'             | region of the world where the event took place                                                           |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'country'            | self-explanatory                                                                                         |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'admin1'             | national administrative regions where the event took place (admin1 being the largest)                    |
|                      |                                                                                                          |
| 'admin2'             |                                                                                                          |
|                      |                                                                                                          |
| 'admin3'             |                                                                                                          |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'location'           | self-explanatory                                                                                         |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'latitude'           | self-explanatory                                                                                         |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'longitude'          | self-explanatory                                                                                         |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'geo_precision'      | numeric code for level of precison of event location                                                     |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'source'             | source of event report                                                                                   |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'source_scale'       | scale (e.g. local, international) of the source                                                          |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'notes'              | description of the event                                                                                 |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'fatalities'         | number of reported fatalities in the event                                                               |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'tags'               | targets (type and estimate counts)                                                                       |
+----------------------+----------------------------------------------------------------------------------------------------------+
| 'time'               | Unix Timestamp of event                                                                                  |
+----------------------+----------------------------------------------------------------------------------------------------------+

</details>

## Myammar's Administrative Boundary data

**Background:**

-   admin 0: country boundary (full overview)

-   1: state/region

-   2: district-level

-   3: township (greatest resolution)

**Approach:**

1.  In the working file directory, group the shapefiles according to admin classification (admin0, admin1, etc.)

2.  Load

3.  Check structure or whatever

::: panel-tabset
### admin 0

```{r}
the1_mymr_shp4326_admin0 <- st_read(dsn = 'data/gadm41_MMR_0/', layer = 'gadm41_MMR_0')
```

```{r}
str(the1_mymr_shp4326_admin0)

# geometry: "XY" "MULTIPOLYGON"
```

```{r}
plot(the1_mymr_shp4326_admin0$geometry)
```

### admin 1

```{r}
the1_mymr_shp4326_admin1 <- st_read(dsn = 'data/gadm41_MMR_1/', layer = 'gadm41_MMR_1')
```

```{r}
str(the1_mymr_shp4326_admin1)
```

```{r}
plot(the1_mymr_shp4326_admin1$geometry)
```

### admin 2

```{r}
the1_mymr_shp4326_admin2 <- st_read(dsn = 'data/gadm41_MMR_2/', layer = 'gadm41_MMR_2')
```

```{r}
str(the1_mymr_shp4326_admin2)
```

```{r}
plot(the1_mymr_shp4326_admin2$geometry)
```

### admin 3

```{r}
the1_mymr_shp4326_admin3 <- st_read(dsn = 'data/gadm41_MMR_3/', layer = 'gadm41_MMR_3')
```

```{r}
str(the1_mymr_shp4326_admin3)
```

```{r}
plot(the1_mymr_shp4326_admin3$geometry)
```
:::

### Deciding on which Boundary

To balance spatial detail and computational efficiency, I think using `the1_mymr_shp4326_admin1` (state/region) boundary data will be the best fit. While looking at district-/township-level provides fine granularity for analysis, for potato machines like my macbook, `the1_mymr_shp4326_admin1` provides the best compromise.

```{r}
tmap_mode('plot')
```

::: panel-tabset
#### admin1

```{r}
qtm(the1_mymr_shp4326_admin1)
```

#### admin3

```{r}
qtm(the1_mymr_shp4326_admin3)
```
:::

```{r}
summary(the1_mymr_shp4326_admin1)
```

**Transform for more precise analysis**

```{r}
st_transform(the1_mymr_shp4326_admin1, crs = 32646) # WGS 84 / UTM zone 46N
```

# Data Wrangling

In this section, I will be doing basic data cleaning and formatting across the ACLED columns so that I can derive appropriate variables needed for analysis.

**Inspecting ACLED**

```{r}
head(acled_sf, n=5)
```

**Check ACLED structure**

```{r}
str(acled_sf)
```

**Check `event_type` values**

```{r}
unique(acled_sf$event_type)
```

**Check for Missing Data**

```{r}
sum(is.na(acled_sf))
```

Clearly, this doesn't provide much insight to the state of the ACLED **columns**. Hence I consulted Uncle Google for alternatives:

-   base R

```{r}
acled_sf %>% 
  summarise(across(everything(), ~ sum(is.na(.)))) %>% 
  
  # pivot_longer() to transpose the table for a more compact output
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") %>%
  
  # filter() to rid off ACLED columns with no missing data.
  filter(missing_count > 0)
```

> '`actor`'-related columns: possible to have 1 or more actors responsible
>
> '`civilian_targeting`': boolean-like variable — null for no civilian targeting
>
> '`admin3`': location identifiers are still discernible at least 1 of the `admin` columns are filled
>
> '`tags`': supplementary target description information

-   `skim()` from `skimr` package

```{r}
skim(acled_sf) %>%
  tibble::as_tibble()
```

**Checking for duplicates in ACLED's unique identifier `event_id_cnty`**

-   base R

```{r}
sum(duplicated(acled_sf['event_id_cnty']))
```

## Wrangling

#### **Reference system**

```{r}
st_crs(acled_sf)
```

```{r}
acled_sf <- acled_sf %>% 
  st_as_sf(coords = c('longitude','latitude'), crs = 4326) %>% 
  st_transform(crs = 32646)
st_crs(acled_sf)
```

#### **Date formatting**

-   `lubridate` package

    -   `as_datetime()`: converts UNIX time-stamps into workable date-time object

    -   `quarter()`: extracts quarter info

```{r}
# to-do: assign variable; mix in geospatial data too
acled_sf_prepped <- acled_sf %>% 
  group_by(event_id_cnty) %>%
  arrange(timestamp) %>% 
  mutate(
    # convert UNIX timestamp into datetime object
    datetime = as_datetime(timestamp, tz = 'Asia/Yangon'),
    
    # extract quarter info
    quarter_num = quarter(datetime),
    
    # concatenate
    quarter_period = paste(year,'Q', quarter_num, sep = '')
  )
```

> Note: Unix timestamps count the number of seconds since the Unix epoch (January 1st, 1970 at UTC)

# `event_type` Exploratory Analysis by Quarter

## Battles

## Explosions/Remote violence

## Protests

## Riots

## Strategic developments

## Violence against civilians

For other data export instructions, visit the link [here](https://acleddata.com/data-export-tool/).
